Increased perceptions of autonomy through choice fail to enhance motor skill retention

### 2021-11-03 docs/r1_submission/r1_manuscript.Rmd
update: include revisions based on reviewer comments

csl               : "../apa.csl"
documentclass     : "apa7"
classoption       : "man" # man for manuscript & jou for journal
output            : papaja::apa6_pdf
header-includes   :
  - \usepackage{pdflscape}
  - \pagewiselinenumbers
  - \raggedbottom
  - \leftheader{\textbf{THIS PREPRINT HAS BEEN SUBMITTED FOR PEER-REVIEW}}
  - \rightheader{\textbf{THIS PREPRINT HAS BEEN SUBMITTED FOR PEER-REVIEW}}
  #- \hypersetup{draft} # comment out after but right now will prevent error message about some link spanning multiple pages
---

@@

Over the years, this manipulation has been described using a variety of names (e.g., self- and/or learner-controlled; -regulated; -directed; -selected), but more recently it has been subsumed by autonomy-support. In fact, autonomy-support is considered such a robust learning variable that it is one of three key pillars in the recently proposed "OPTIMAL" theory of motor learning [@wulf2016]. @wulf2016 argued that providing learners with opportunities for choice---considered an autonomy-supportive practice manipulation---can facilitate motor performance and learning by enhancing learner's expectancies for success, and by allowing the learner to maintain their attentional focus on the task by reducing the need for self-regulatory activity [@abdollahipour2017; @chua2018; @wulf2018]. In other words, these psychological and attentional benefits, and concomitant increases in performance and learning are a by-product of exercising choice during practice. Overall, autonomy-support is seen as a means to efficient goal-action coupling [@wulf2016] and also links the "OPTIMAL" theory with Self-Determination theory [@deci2000].

Despite its prominent role within the "OPTIMAL" theory of motor learning, there are numerous reasons to doubt the importance of autonomy-support through the provision of choice during practice. First, if the benefits are the result of having opportunities for choice then learning differences should not emerge between different self-controlled groups. However, in experiments where different groups of participants have choice over their feedback schedule, such learning differences have been found, but not limited to, when this choice is made *after* rather than before a performance attempt [@chiviacowsky2005; @carter2014], when different criteria for success are provided to participants [@chiviacowsky2012], and when the absolute number of feedback choice opportunities are limited compared to unlimited at the outset of practice [@hansen2011; @patterson2011]. Second, there has been little-to-no support for the notion that practicing in a self-controlled group is perceived as more autonomy-supportive than being in a yoked group. @ste-marie2013, for example, had participants in the self-controlled and yoked groups complete the perceived choice subscale of the Intrinsic Motivation Inventory and failed to find the expected effect of higher self-reported scores during practice in the self-controlled group. Similar findings have been reported by others [@barros2019; @carter2017a; @mckay2020b]. However, @mckay2020a recently found that practicing in a self-controlled group was perceived as more autonomy-supportive than practicing the same task in a yoked group. This increased perceived autonomy-support did not, however, translate into enhanced learning compared to the yoked group. Although the available literature does not provide substantial evidence that self-controlled practice is autonomy-supportive, this does not necessarily mean such a manipulation is not autonomy-supportive [@altman1995]. For instance, this effect may in fact be quite small and require much larger sample sizes to detect than those commonly used in motor learning experiments [@lohse2016].

There are at least two other methodological issues that warrant consideration. The first of these is that in self-controlled motor learning experiments participants in the self-controlled group are usually given choice over a single component (e.g., feedback or when to watch a modeled demonstration) of their practice and participants in the yoked group are not given choice over this component. However, within the context of practice itself there are many other opportunities for choice that participants may explore, independent of their assigned group. Target-based tasks such as basketball free throws [e.g., @aiken2012] or bean-bag tossing [@grand2015] are a popular choice in motor learning experiments and have been used in self-controlled learning experiments. While a participant in the yoked group may not be permitted choice over their feedback schedule (or some other practice variable) in these experiments, this does not preclude them from opportunities for choice---and thus autonomy-support---when deciding to try different throwing techniques, speeds, and/or release points. Thus, labeling yoked groups as being devoid of choice opportunities may be a misnomer. The other consideration relates to the instructions that are provided to participants in the yoked group. In the context of feedback[^1], these participants are typically informed that during practice they may or may not receive feedback after a given trial [e.g., @chiviacowsky2002; @patterson2010]. This means that participants in yoked groups are not even aware that they have been denied an opportunity for choice, nor that their feedback schedule was created by another participant who was given choice over when feedback was or was not provided. Either of these in isolation, or both simultaneously could contribute to the consistent finding that participants in self-controlled and yoked groups report similar perceived autonomy scores when asked about their opportunities for choices with respect to the motor task [@ste-marie2013] or about their practice environment in general [@barros2019; @carter2017a; @mckay2020b].

[^1]: While feedback is used in this example, this issue surrounding instructions is also relevant to other practice variables commonly used in self-controlled learning experiments.

>>>>

documentclass     : "apa7"
classoption       : "man, donotrepeattitle" # man for manuscript & jou for journal
output            : papaja::apa6_pdf
header-includes   :
  - \usepackage{pdflscape}
  - \pagewiselinenumbers
  - \raggedbottom
  - \leftheader{\textbf{THIS PREPRINT HAS BEEN SUBMITTED FOR PEER-REVIEW}}
  - \rightheader{\textbf{THIS PREPRINT HAS BEEN SUBMITTED FOR PEER-REVIEW}}
  #- \hypersetup{draft} # comment out after but right now will prevent error message about some link spanning multiple pages
---

@@

Over the years, this manipulation has been described using a variety of names (e.g., self- and/or learner-controlled; -regulated; -directed; -selected), but more recently it has been subsumed by autonomy-support. \textcolor{red}{Within their ``OPTIMAL'' theory of motor learning,} @wulf2016 \textcolor{red}{argued that providing learners with opportunities for choice creates an autonomy-supportive practice environment, which facilitates motor performance and learning. Specifically, the authors predict that autonomy-support facilitates performance by enhancing expectancies (Prediction 3, p. 1404), that enhanced expectancies and autonomy support contribute to efficient goal-action coupling by readying the motor system for task execution (Prediction 2, p. 1404), and that enhanced expectancies and autonomy support facilitate motor learning by making dopamine available for memory consolidation and neural pathway development (Prediction 7, p. 1404). In other words, these psychological benefits of increased perceptions of autonomy and competence, and the resulting increases in performance and learning are a by-product of having choice itself. Overall,} @wulf2016's \textcolor{red}{``OPTIMAL'' theory of motor learning provides a motivational explanation for the learning advantages of self-controlled practice conditions over yoked practice conditions.}[^1]

[^1]: \textcolor{red}{It should be noted that other researchers have instead presented an information-processing explanation for the self-controlled learning advantages} [see @ste-marie2019 for a recent discussion of the motivational and information-processing explanations]\textcolor{red}{. We acknowledge this view here; however, unlike previous experiments} [e.g., @barros2019; @carter2014; @carter2017a; @carter2017b; @couvillion2020; @woodard2020] \textcolor{red}{the current experiment was not designed to test between explanations. We focused on the motivational explanation as this view has garnered more attention due the ``OPTIMAL'' theory of motor learning.}

\textcolor{red}{Despite its prominent role as a robust and generalizable learning variable in the ``OPTIMAL'' theory (p. 1393), there is considerable ambiguity surrounding whether the provision of choice is in fact an autonomy-supportive manipulation. First, the notion that practicing in a self-controlled group is actually more autonomy-supportive than a yoked group has primarily been assumed} [e.g., @abdollahipour2017; @chua2018; @lewthwaite2015; @wulf2018] \textcolor{red}{rather than supported empirically. Second, when researchers have included measures related to perceptions of autonomy the data is mixed. For example,} @ste-marie2013 \textcolor{red}{did not find the expected effect of higher perceptions of autonomy during practice in a self-controlled group as compared to a yoked group. Similar outcomes have been reported by others} [e.g., @barros2019; @carter2017a; @mckay2020b]. \textcolor{red}{In contrast,} @mckay2020a \textcolor{red}{recently found that practicing in a self-controlled group was perceived as more autonomy-supportive than practicing the same task in a yoked group. However, the higher perceived autonomy scores did not translate into enhanced motor performance or learning in the self-controlled group as predicted in the ``OPTIMAL'' theory. Although the majority of experiments that included a measure related to perceived autonomy reported no group differences, the ``absence of evidence is not evidence of absence''} [@altman1995]\textcolor{red}{. Thus, self-controlled practice conditions could be an autonomy-supportive manipulation but such an effect may actually be quite small and require much larger sample sizes to detect than those used in previous experiments} [e.g., @barros2019; @ste-marie2013] \textcolor{red}{and in motor learning experiments in general} [see @lohse2016 for a discussion]\textcolor{red}{. This argument of underpowered experimental designs is supported by the results of} @mckay2020a \textcolor{red}{as these authors had one of the largest sample sizes to date in the self-controlled literature.}

There are at least two other methodological issues that warrant consideration. In self-controlled motor learning experiments participants in the self-controlled group are usually given choice over a single component (e.g., feedback or when to watch a modeled demonstration) of their practice and participants in the yoked group are not given choice over this component. However, within the context of practice itself there are other opportunities for choice that participants may explore, independent of their assigned group. Past self-controlled learning experiments have used a variety of motor tasks, including but not limited to basketball free throws [e.g., @aiken2012] and bean-bag tossing [e.g., @grand2015]. Although participants in a yoked group may not have choice over their feedback schedule (or some other practice variable) while learning such tasks, this does not preclude them from opportunities for choice---and thus autonomy-support---when choosing to try different throwing techniques, speeds, and/or release points. Thus, labeling yoked groups as being devoid of choice opportunities may be a misnomer. The other issue relates to the instructions that are provided to participants in a yoked group. In the context of feedback[^2], these participants are typically informed that during practice they may or may not receive feedback after a given trial [e.g., @chiviacowsky2002; @patterson2010]. This means that participants in yoked groups are not even aware that they have been denied an opportunity for choice, or that their feedback schedule was created by another participant who had choice over when feedback was or was not provided. These two methodological issues, either in isolation or simultaneously, may have contributed to the consistent finding in past research with self-controlled and yoked participants self-reporting numerically similar perceptions of autonomy when asked about opportunities for choice with respect to the motor task [e.g., @ste-marie2013] or about their practice environment in general [e.g., @barros2019; @carter2017a; @mckay2020b].

[^2]: While feedback is used in this example, this issue surrounding instructions is also relevant to other practice variables commonly used in self-controlled learning experiments.
